{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-04-28 paper.tex
first pass on introduction

\begin{abstract}
We present a new method conceived to ease the inclusion of hadronic
observables into parton distribution function (PDF) fits. Our
technique relies on the the same principles adopted by the existing
fast interfaces but implements a set of improvements aimed at
optimizing the performance in the context of PDF fits where many
iterations and many predictions for each iteration are usually
required. The main novelties are the precomputation of the PDF and
$\alpha_s$ evolution and the optimisation of the numerical convolution
between hard cross sections and PDFs. We demonstrate that our technique
can lead to a substantial speed improvement as compared to the
existing interfaces.
\end{abstract}

@@

\begin{small}
\noindent
{\em Manuscript Title:} {\tt APFELgrid}: a fast interface to hadronic observables for
  fits of parton densities                                      \\
{\em Authors:} V.~Bertone, S.~Carrazza, N.P.~Hartland                                               \\
{\em Program Title:} {\tt APFELgrid}                                          \\
{\em Journal Reference:}                                      \\
  %Leave blank, supplied by Elsevier.
{\em Catalogue identifier:}                                   \\
  %Leave blank, supplied by Elsevier.
{\em Licensing provisions:}                                   \\
  %enter "none" if CPC non-profit use license is sufficient.
{\em Programming language:}                                   \\
{\em Computer:}                                               \\
  %Computer(s) for which program has been designed.
{\em Operating system:}                                       \\
  %Operating system(s) for which program has been designed.
{\em RAM:} bytes                                              \\
  %RAM in bytes required to execute program with typical data.
{\em Number of processors used:}                              \\
  %If more than one processor.
{\em Supplementary material:}                                 \\
  % Fill in if necessary, otherwise leave out.
{\em Keywords:} Keyword one, Keyword two, Keyword three, etc.  \\
  % Please give some freely chosen keywords that we can use in a
  % cumulative keyword index.
{\em Classification:}                                         \\
  %Classify using CPC Program Library Subject Index, see (
  % http://cpc.cs.qub.ac.uk/subjectIndex/SUBJECT_index.html)
  %e.g. 4.4 Feynman diagrams, 5 Computer Algebra.
{\em External routines/libraries:}                            \\
  % Fill in if necessary, otherwise leave out.
{\em Subprograms used:}                                       \\
  %Fill in if necessary, otherwise leave out.
{\em Catalogue identifier of previous version:}*              \\
  %Only required for a New Version summary, otherwise leave out.
{\em Journal reference of previous version:}*                  \\
  %Only required for a New Version summary, otherwise leave out.
{\em Does the new version supersede the previous version?:}*   \\
  %Only required for a New Version summary, otherwise leave out.
{\em Nature of problem:}\\
  %Describe the nature of the problem here.
   \\
{\em Solution method:}\\
  %Describe the method solution here.
   \\
{\em Reasons for the new version:}*\\
  %Only required for a New Version summary, otherwise leave out.
   \\
{\em Summary of revisions:}*\\
  %Only required for a New Version summary, otherwise leave out.
   \\
{\em Restrictions:}\\
  %Describe any restrictions on the complexity of the problem here.
   \\
{\em Unusual features:}\\
  %Describe any unusual features of the program/problem here.
   \\
{\em Additional comments:}\\
  %Provide any additional comments here.
   \\
{\em Running time:}\\
  %Give an indication of the typical running time here.
   \\
* Items marked with an asterisk are only required for new versions
of programs previously published in the CPC Program Library.\\
\end{small}

@@

\tableofcontents

@@

The need for fast and accurate cross-section predictions in the Large
Hadron Collider (LHC) era is becoming increasingly important. In fact,
the outstanding amount of precise data produced by the LHC experiments
is bound to provide a very stringent constraint on the proton parton
distribution functions (PDFs) on the condition that as many
measurements as possible are included in PDF fits with the highest
accuracy possible. However, the direct use of most of the codes that
provide predictions for non-inclusive hadron-collider observables in a
PDF fit, which typically requires a large number of iterations, is not
viable because the time required to obtain accurate predictions is
usually of the order of a few hours or more. In order to overcome this
limitation, the typical strategy adopted relies on the precomputation
of the partonic hard cross sections in such a way that the numerical
convolution with any set of PDFs can be quickly performed {\it a
  posteriori} by means of interpolation techniques.

Examples of the implementation of this strategy are {\tt
  APPLgrid}~\cite{Carli:2010rw} and {\tt
  FastNLO}~\cite{Wobisch:2011ij}. For the computation of the hard
cross sections, these packages rely on external codes to which they
are interfaced, more or less directly, by means of a suite of
functions whose scope is that of filling up PDF- and
$\alpha_s$-independent look-up tables of cross-section weights that can
subsequently be combined with an arbitrary set of PDFs. Examples of
codes that have been directly interfaced to {\tt APPLgrid} and/or to
{\tt FastNLO} are {\tt MCFM}~\cite{Campbell:2010ff} and {\tt
  NLOJet++}~\cite{Nagy:2003tz}. More recently, dedicated interfaces to
automated general-purpose Monte Carlo (MC) event generators have been
developed. In particular, the {\tt aMCfast} and {\tt MCgrid} codes
have been specifically designed to extract the relevant information
from the {\tt MadGraph5\_aMC@NLO}~\cite{Alwall:2014hca} and {\tt
  SHERPA}~\cite{Gleisberg:2008ta} event generators, respectively, and
fill up look-up interpolation tables in the {\tt APPLgrid} and/or {\tt
  FastNLO} format.

While these tools have proven to be extremely useful for the
extraction of parton densities, the great amount of experimental data
that has become available thanks to the LHC and that are or are going
to be included into the current PDF analyses is stretching the
capabilities of the general strategy as well as its
implementations. In fact, presently a typical global PDF fit includes
thousands of hadronic data points whose predictions have to be
computed thousands of times during the minimisation process. As a
consequence, the ``standard'' tools, $i.e.$ {\tt APPLgrid} and {\tt
  FastNLO}, might not be fast enough to meet the needs of the modern
PDF fitters. For this reason we believe that a high-performance tool
tailored for PDF analyses is becoming increasingly desirable.

The so-called {\tt FastKernel} method, developed in the framework of
the NNPDF global analyses~\cite{Ball:2010de}, was conceived to address
this problem. The basic principle of this method is the maximisation
of the precomputation with the consequent minimisation of the amount
of operations needed to be performed during the PDF fit. In
particular, the {\tt FastKernel} method relies on the precomputation
of the partonic hard cross sections and the DGLAP evolution kernels
and their combination into a single look-up table, which we call {\tt
  FKtable}, in such a way that the prediction for a given hadronic
observable can be obtained by a single numerical convolution between
the respective {\tt FKtable} and the initial scale PDFs, which are
the quantities that are usually fitted to data.

In this paper we present a public implementation of the {\tt
  FastKernel} method in which the partonic hard cross sections are
taken from any given {\tt APPLgrid} look-up table while the DGLAP
evolution kernels are provided by the {\tt APFEL}
code~\cite{Bertone:2013vaa}. We named the resulting code {\tt
  APFELgrid}.

>>>>

\begin{abstract}
We present a new software package designed to aid the inclusion of hadronic
observables into Parton Distribution Function (PDF) fits. The {\tt APFELgrid} package
converts interpolated weight tables provided by {\tt APPLgrid} files into a more efficient
format for PDF fitting by the pre-computation and combination with PDF and $\alpha_s$ evolution factors provided
by {\tt APFEL}. This combination significantly reduces the number of operations required to perform the calculation
of hadronic observables in PDF fits and simplifies the structure of the calculation into
a readily optimised scalar product. We demonstrate that our technique
can lead to a substantial speed improvement as compared to the
existing interfaces without any reduction in numerical accuracy.

%We present a new method conceived to ease the inclusion of hadronic
%observables into parton distribution function (PDF) fits. Our
%technique relies on the the same principles adopted by the existing
%fast interfaces but implements a set of improvements aimed at
%optimizing the performance in the context of PDF fits where many
%iterations and many predictions for each iteration are usually
%required. The main novelties are the precomputation of the PDF and
%$\alpha_s$ evolution and the optimisation of the numerical convolution
%between hard cross sections and PDFs. We demonstrate that our technique
%can lead to a substantial speed improvement as compared to the
%existing interfaces.
\end{abstract}

@@

\begin{small}
\noindent
{\em Manuscript Title:} {\tt APFELgrid}: a fast interface to hadronic observables for
  fits of parton densities                                      \\
{\em Authors:} V.~Bertone, S.~Carrazza, N.P.~Hartland                                               \\
{\em Program Title:} {\tt APFELgrid}                                          \\
{\em Journal Reference:}                                      \\
  %Leave blank, supplied by Elsevier.
{\em Catalogue identifier:}                                   \\
  %Leave blank, supplied by Elsevier.
{\em Licensing provisions:}   MIT license                                \\
{\em Programming language:}  C++                                 \\
{\em Computer:} PC/Mac                                               \\
  %Computer(s) for which program has been designed.
{\em Operating system:} MacOS/Linux                                       \\
  %Operating system(s) for which program has been designed.
{\em RAM:} varying                                              \\
  %RAM in bytes required to execute program with typical data.
{\em Keywords:} QCD, PDF\\
  % Please give some freely chosen keywords that we can use in a
  % cumulative keyword index.
{\em Classification:}  11.1 General, High Energy Physics and Computing                                       \\
  %Classify using CPC Program Library Subject Index, see (
  % http://cpc.cs.qub.ac.uk/subjectIndex/SUBJECT_index.html)
  %e.g. 4.4 Feynman diagrams, 5 Computer Algebra.
{\em External routines/libraries:}  {\tt APPLgrid}, {\tt APFEL}                          \\
  % Fill in if necessary, otherwise leave out.
{\em Nature of problem:}\\
 Fast computation of hadronic observables under the variation of parton distribution functions.
   \\
{\em Solution method:}\\
  Combination of interpolated weight grids from {\tt APPLgrid} and evolution factors from {\tt APFEL} into efficient {\tt FastKernel} tables.\\
{\em Running time:} varying\\
   \\
\end{small}

@@

%%\tableofcontents

@@

The ability to perform fast and accurate cross-section predictions for measurements at
the Large Hadron Collider (LHC) is a vital requirement for the most precise PDF determinations. LHC measurements have a unique
capacity to shed light upon the dynamics of the proton and provide stringent constraints upon 
the proton PDF. In order to make full use of current and future LHC data, efficient strategies
for the computation of predictions for these observables must be deployed. Software packages performing predictions 
for non-inclusive hadron-collider observables cannot be easily deployed in a PDF fit requiring rapid iteration due the time required to obtain accurate results (usually of the order of a few hours or more
per data point). In order to overcome this limitation, the typical strategy adopted for fast cross-section prediction relies on the precomputation
of the partonic hard cross sections in such a way that the numerical convolution with any set of PDFs can be quickly performed by means of interpolation techniques.

This interpolation strategy is implemented in the {\tt
  APPLgrid}~\cite{Carli:2010rw} and {\tt
  FastNLO}~\cite{Wobisch:2011ij} projects. For the computation of the hard
cross sections, these packages rely on external codes to which they
are interfaced by means of a suite of
functions allowing for the filling of PDF- and
$\alpha_s$-independent look-up tables of cross-section weights. Codes
such as {\tt MCFM}~\cite{Campbell:2010ff} and {\tt NLOJet++}~\cite{Nagy:2003tz} have been interfaced directly to 
 {\tt APPLgrid}/{\tt FastNLO} and more recently dedicated interfaces to
automated general-purpose Monte Carlo (MC) event generators have been
developed. The {\tt aMCfast} and {\tt MCgrid} codes
have been specifically designed to extract the relevant information
from the {\tt MadGraph5\_aMC@NLO}~\cite{Alwall:2014hca} and {\tt
  SHERPA}~\cite{Gleisberg:2008ta} event generators respectively.

While these tools have proven to be extremely useful in the
extraction of parton densities, the volume of experimental data
made available by LHC collaborations for use in PDF fits is already
stretching the capabilities of typical fitting technology. Presently a typical global PDF fit includes
thousands of hadronic data points for which predictions have to be
computed thousands of times during the minimisation process. As a
consequence the ``standard'' tools, $i.e.$ {\tt APPLgrid} and {\tt
  FastNLO}, might not be fast enough to meet the needs of modern
PDF fitters. For this reason a high-performance tool tailored specifically 
for PDF analyses is becoming increasingly important.

The {\tt FastKernel} method, developed in the framework of
the NNPDF global analyses~\cite{Ball:2010de}, was developed to address
this problem. This method differs from the standard procedure \`{a} la {\tt APPLgrid} or {\tt fastNLO} in that it
maximises the amount of information that is precomputed prior to fitting such as to minimise the amount
of operations to be performed during the fit. Specifically, the {\tt FastKernel} method relies on the precomputation
of the partonic hard cross sections and DGLAP evolution kernels
and their combination into a single look-up table, here called a {\tt FastKernel} ({\tt FK}) table, in such a way that the prediction for a given hadronic
observable can be obtained by a simple matrix product between
the respective {\tt FK} table and the initial scale PDFs.

In this paper we present the {\tt APFELgrid} package, a public implementation of the {\tt
  FastKernel} method in which the hard partonic cross sections provided in an {\tt APPLgrid} look-up table are combined with the DGLAP
evolution kernels provided by the {\tt APFEL}
package~\cite{Bertone:2013vaa}.
